Autonomous Oceanographic Sampling Designs using Excursion Sets for Multivariate Gaussian random fields

### 2020-06-04 main_new.tex
updates

	In the particular case where $\gp$ is a multivariate Gaussian random field 
	%and denoting the mean and covariance matrix of $\gp[\bm{u}]$ by $\meanUU\in\mathbb{R}^{p\times r}$ and $\covUU$, respectively, 
	we have 
	\begin{align*}
	\jointExcuProb = \mathcal{N}_{\no r}(T^r; \meanUU, \covUU),
	\end{align*}
	where $\mathcal{N}_{\no r}(\cdot ; \meanUU, \covUU)$ is the Gaussian measure on $\mathbb{R}^{\no r}$ with mean $\meanUU$ and covariance matrix $\covUU$, respectively defined blockwise by  
	%where both are defined blockwise by 
	\begin{align*}
	\meanUU&=\begin{pmatrix}\mu(u^{(1)})\\ \vdots\\ \mu(u^{(r)})\end{pmatrix}
	\in \mathbb{R}^{\no r}, \\
	\text{and } \covUU &= \begin{pmatrix}
	\cov(\gp[u^{(1)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(1)}],
	\gp[u^{(r)}])\\
	\vdots & & \vdots\\
	\cov(\gp[u^{(r)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(r)}],
	\gp[u^{(r)}])\\
	\end{pmatrix}\in \mathbb{R}^{pr\times pr},
	\end{align*}
	each of the $r\times r$ blocks of the latter matrix being itself a (cross-)covariance matrix of dimension $\no \times \no$. 
	%dimensional Gaussian random vector. 
	Assuming further that $\covUU$ is non-singular, the probability of interest can be formulated in terms of the $\no r$-dimensional Gaussian probability density function 
	$\varphi_{\no  r}(\cdot;~\meanUU, \covUU)$ as 
	%$\varphi_{\no \times r}$ as
	%$\bm{u}\in D^r$ such that $\gp[\bm{u}]$ be non-degenerate, the integrand can be expressed in terms of the $r \times \no$-dimensional Gaussian distribution, namely 
	%cumulative distribution function, namely 
	\begin{equation*}
	\begin{split}
	&\jointExcuProb
	=
	\int_{T^r} \varphi_{\no  r}\left(\bm{v};
	~\meanUU, \covUU\right) 
	\mathrm{d}\bm{v},
	\end{split}
	\end{equation*}
	In the particular orthant case with $\T=(-\infty, t_1] \times \dots \times (-\infty, t_{r}]$,
	the latter probability directly writes in terms of the multivariate Gaussian
	cumulative distribution, % with mean $\meanUU$ and covariance matrix $\covUU$,
	this time without requiring $\covUU$ %the latter 
	to be non-singular: 
	\begin{equation*}
	\begin{split}
	\jointExcuProb 
	&=
	\varPhi_{\no r}\left(\bm{t};~\meanUU, \covUU\right),
	\end{split}
	\end{equation*}
	where we have used the notations 
	$t=(t_1,\dots,t_{\no}%,..., t_1, ..., t_p
	)\in\mathbb{R}^{\no}$, $1_{r}=(1,\dots,1)\in \R^{r}$, and 
	$\bm{t}=1_{r}\otimes \bm{t}=(t_1,\dots,t_{\no},\dots,t_1,\dots,t_{\no}) 
	\in \R^{\no r}$.
\end{propo}

@@

% In this section, 
We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity. \kc{What is the source of this data?}
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.pdf}
  \caption{Density contour plots with different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. X-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean $\mu_1=t_1=5^o C$, and
salinity $\mu_2=t_2=30$ mg/l, and the EP is $p=P(Z_1 < 5,Z_2 < 30)$.
The displayed densities have unit standard deviations for both
temperature and salinity, but we also study the effect of doubling the
standard deviations.

Table \ref{tab:sim_rhoab} shows the initial EPs and the associated
Bernoulli variance (second row) for the examples indicated in Fig.
\ref{illus_bivarDens}. The EPs increase with the correlation as there
is a strong tendency to have concurrently low temperature and
salinity. The Bernoulli variance is similarly large for high
correlations. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

@@

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types, and for a design with
temperature measurements alone. Both data are modeled by $(Y_1,Y_2)^t=(Z_1,Z_2)^t+N(0,0.5^2I_2)$, while
$Y_2=Z_2+N(0,0.5^2)$ when only temperature is measured.  For this
illustration, Table \ref{tab:sim_rhoab} shows that the expected
Bernoulli variance gets lower with larger standard deviations
$\sigma_1$ and $\sigma_2$ (right columns). The reduction of Bernoulli
variance is largest for the cases with high correlation
$\gamma$. Albeit smaller, there is also uncertainty reduction when
only temperature is measured (bottom row), especially when temperature
and salinity are highly correlated. When correlation is low
($\gamma=0.2$), there is little information about salinity in the
temperature data, and therefore less uncertainty reduction. In an
application with fresh cold water from a river source, the temperature
and salinity variables will not only be interdependent, but will also
likely show dependence in the spatial dimension. This in turn will
impact the design criteria when we evaluate the information measure by
integrating over several locations (Section \ref{sec:simulations}).

@@

An adaptive AUV survey is split into many stages. The selection on where to sample is made with restriction to the triangular waypoint graph and defined by its nearest grid nodes.

@@

At the next stage, we choose where measurements $\by_{d^{j,s}}$ should be gathered with
design $d^{j,s}$. For the triangularized waypoint graph that we use in our AUV application, there are six possible directions for $j$, and we must evaluate the design criterion for all these. 
The goal of the sampling design is to minimize the expected uncertainty of random sets, and specifically the EIBV for our application. We hence focus on choosing designs for data modeled by equation \eqref{eq:meas}, to conduct efficient sequential uncertainty reduction according to the EIBV criterion.
Notably, compared with the static design $d$ in equation \eqref{eq:eibv}, the optimal sequential design not only considers the best current AUV
design grid point, but also how the data gathered at this node could inform future sampling at successive stages.

@@

The simplest heuristic for adaptive sampling is to choose the next
sampling location based on current probabilities. One reduces the largest levels of uncertainty by selecting the design $j$ with $P(Z(\x_j) \in T)$ closest to $0.5$.
For our situation with temperature and salinity variables, this {\it{naive}} strategy selects the stage $s$ sampling design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j} |P(Z_1(\x_j) \leq t_1, Z_2(\x_j) \leq t_2 | \mathcal{Y}^{s-1})-0.5|, \nonumber
\end{eqnarray}
where $j$ is indexing all neighboring locations on the waypoint graph.

This strategy does not account for the uncertainty in the GP variables, but is instead applicable only if one design (node) has EPs closer to
$0.5$ (see Table \ref{tab:sim_rhoab}, line two). Additionally, this strategy does not
account for spatial correlation. This strategy lacks memory of where
it has been and where the uncertainty has been reduced, and is therefore
susceptible to local minima.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more stage of measurements. This selection
strategy is based on expected uncertainty reduction, but it is heuristic because there is no anticipation of what the subsequent designs might
offer beyond the first stage.

>>>>

	In the particular case where $\gp$ is a multivariate Gaussian random field 
	%and denoting the mean and covariance matrix of $\gp[\bm{u}]$ by $\meanUU\in\mathbb{R}^{p\times r}$ and $\covUU$, respectively, 
	we have 
	\begin{align*}
	\jointExcuProb = \mathcal{N}_{\no r}(T^r; \meanUU, \covUU),
	\end{align*}
	where $\mathcal{N}_{\no r}(\cdot ; \meanUU, \covUU)$ is the Gaussian measure on $\mathbb{R}^{\no r}$ with mean $\meanUU$ and covariance matrix $\covUU$, respectively defined blockwise by  
	%where both are defined blockwise by 
	\begin{align*}
	\meanUU&=\begin{pmatrix}\mu(u^{(1)})\\ \vdots\\ \mu(u^{(r)})\end{pmatrix}
	\in \mathbb{R}^{\no r}, \\
	\text{and } \covUU &= \begin{pmatrix}
	\cov(\gp[u^{(1)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(1)}],
	\gp[u^{(r)}])\\
	\vdots & & \vdots\\
	\cov(\gp[u^{(r)}], \gp[u^{(1)}]) & \dots & \cov(\gp[u^{(r)}],
	\gp[u^{(r)}])\\
	\end{pmatrix}\in \mathbb{R}^{pr\times pr},
	\end{align*}
	each of the $r\times r$ blocks of the latter matrix being itself a (cross-)covariance matrix of dimension $\no \times \no$. 
	%dimensional Gaussian random vector. 
	Assuming further that $\covUU$ is non-singular, the probability of interest can be formulated in terms of the $\no r$-dimensional Gaussian probability density function 
	$\varphi_{\no  r}(\cdot;~\meanUU, \covUU)$ as 
	%$\varphi_{\no \times r}$ as
	%$\bm{u}\in D^r$ such that $\gp[\bm{u}]$ be non-degenerate, the integrand can be expressed in terms of the $r \times \no$-dimensional Gaussian distribution, namely 
	%cumulative distribution function, namely 
	\begin{equation*}
	\begin{split}
	&\jointExcuProb
	=
	\int_{T^r} \varphi_{\no  r}\left(\bm{v};
	~\meanUU, \covUU\right) 
	\mathrm{d}\bm{v},
	\end{split}
	\end{equation*}
	In the particular orthant case with $\T=(-\infty, t_1] \times \dots \times (-\infty, t_{r}]$,
	the latter probability directly writes \kc{What does
          ``writes'' mean here? Connotes?} in terms of the multivariate Gaussian
	cumulative distribution, % with mean $\meanUU$ and covariance matrix $\covUU$,
	this time without requiring $\covUU$ %the latter 
	to be non-singular: 
	\begin{equation*}
	\begin{split}
	\jointExcuProb 
	&=
	\varPhi_{\no r}\left(\bm{t};~\meanUU, \covUU\right),
	\end{split}
	\end{equation*}
	where we have used the notations 
	$t=(t_1,\dots,t_{\no}%,..., t_1, ..., t_p
	)\in\mathbb{R}^{\no}$, $1_{r}=(1,\dots,1)\in \R^{r}$, and 
	$\bm{t}=1_{r}\otimes \bm{t}=(t_1,\dots,t_{\no},\dots,t_1,\dots,t_{\no}) 
	\in \R^{\no r}$.
\end{propo}

@@

% In this section, 
We calculate EPs and the Bernoulli variance for a few bivariate
Gaussian distributions to illustrate the concepts articulated above.
Fig. \ref{illus_bivarDens} shows contour plots of three different
densities with increasing correlation $\gamma$ between temperature and
salinity. \kc{What is the source of this data and this figure,
  i.e. how did you derive this plot?}
\begin{figure}[h!] \centering
  \includegraphics[width=0.99\textwidth]{Figures/illus_bivar.pdf}
  \caption{Density contour plots with different correlations between
    temperature and salinity. The densities have unit variance and the
    thresholds are identical to the mean values $5^o C$ and
    $30 mg/l$. X-axis is temperature and y-axis is salinity.}
\label{illus_bivarDens}
\end{figure}
Here, the thresholds are set equal to the mean $\mu_1=t_1=5^o C$, and
salinity $\mu_2=t_2=30$ mg/l, and the EP is $p=P(Z_1 < 5,Z_2 < 30)$.
The displayed densities have unit standard deviations for both
temperature and salinity, but we also study the effect of doubling the
standard deviations.

Table \ref{tab:sim_rhoab} shows the initial EPs and the associated
Bernoulli variance (second row) for the examples indicated in Fig.
\ref{illus_bivarDens}. The EPs increase with the correlation as there
is a strong tendency to have concurrently low temperature and salinity
\kc{Same question as above. What is the basis for this correlation? Is
  this just an example or derived from actual data in the mixed layer
  in the front?}. The Bernoulli variance is similarly large for high
correlations. EPs and Bernoulli variances are the same for standard
deviation $1$ or $2$, which implies that high variability in
temperature and salinity is not captured in the $p(1-p)$ expression.

@@

Table \ref{tab:sim_rhoab} (bottom two rows) shows results of expected
Bernoulli variance calculations. This is presented for a design
gathering both data types, and for a design with temperature
measurements alone. Both data are modeled by
$(Y_1,Y_2)^t=(Z_1,Z_2)^t+N(0,0.5^2I_2)$, while $Y_2=Z_2+N(0,0.5^2)$
when only temperature is measured \kc{I find this notation somewhat
  confusing. Earlier you had set up t1 and t2 to denote temp and
  salinity respectively. It might help to have \emph{some} notational
  similarity with that; otherwise it becomes mind bending I believe.}.
For this illustration, Table \ref{tab:sim_rhoab} shows that the
expected Bernoulli variance gets lower with larger standard deviations
$\sigma_1$ and $\sigma_2$ (right columns). The reduction of Bernoulli
variance is largest for the cases with high correlation
$\gamma$. Albeit smaller, there is also uncertainty reduction when
only temperature is measured (bottom row), especially when temperature
and salinity are highly correlated. When correlation is low
($\gamma=0.2$), there is little information about salinity in the
temperature data, and therefore less uncertainty reduction. In an
application with fresh cold water from a river source, the temperature
and salinity variables will not only be interdependent, but will also
likely show dependence in the spatial dimension. This in turn will
impact the design criteria when we evaluate the information measure by
integrating over several locations (Section \ref{sec:simulations}).

\kc{I feel this section above, which was present in the earlier
  submitted version too, but to which I did not comment, is
  significant as an example to the reader. But, as noted above, I
  don't understand the provenance of the data. It isn't made up,
  right?}

@@

Our adaptive AUV survey is split into many stages. The selection on
where to sample is constrained by a triangular waypoint graph and
defined by its nearest grid nodes.

@@

At the next stage, we choose where measurements $\by_{d^{j,s}}$ should
be gathered with design $d^{j,s}$. For the triangularized waypoint
graph that we use in our application, there are six possible
directions for $j$, and we must evaluate the design criterion for all
these.  The goal of the sampling design is to minimize the expected
uncertainty of random sets, and specifically the EIBV for our
application. We hence focus on choosing designs for data modeled by
equation \eqref{eq:meas}, to conduct efficient sequential uncertainty
reduction according to the EIBV criterion.  Notably, compared with the
static design $d$ in equation \eqref{eq:eibv}, the optimal sequential
design not only considers the best current design grid point, but also
how the data gathered at this node could inform future sampling at
successive stages.

@@

The simplest heuristic for adaptive sampling is to choose the next
sampling location based on current probabilities. One reduces the
largest levels of uncertainty by selecting the design $j$ with
$P(Z(\x_j) \in T)$ closest to $0.5$.  With temperature and salinity
variables, this {\it{naive}} strategy selects the stage $s$ sampling
design according to:
\begin{eqnarray}\label{critNaive}
    d^{*,s} &=& \mbox{argmin}_{j} |P(Z_1(\x_j) \leq t_1, Z_2(\x_j) \leq t_2 | \mathcal{Y}^{s-1})-0.5|, \nonumber
\end{eqnarray}
where $j$ is indexing all neighboring locations on the waypoint graph.

This strategy does not account for the uncertainty in the GP
variables, but is instead applicable only if one design (node) has EPs
closer to $0.5$ (see Table \ref{tab:sim_rhoab}, line \kc{row?}
two). Additionally, this strategy does not account for spatial
correlation. This strategy lacks memory of where it has been and where
the uncertainty has been reduced, and therefore susceptible to local
minima.

@@

The myopic (greedy) strategy which we present here is optimal if we
imagine taking only one more stage of measurements. This selection
strategy is based on expected uncertainty reduction, but it is a
heuristic because there is no anticipation of what the subsequent
designs might offer beyond the first stage.
